---
title: "gur_file"
format: html
editor: visual
---

### Question 3

```{r}

library(dplyr)
library(tidyverse)
library(ggplot2)
library(tidyr)
library(broom)

x <- election_data_tidy
# election_data_tidy <- x

election_data_tidy <- election_data_tidy |> 
  group_by(cod_mun, anno, party_recoded) |> 
  mutate(
    total_votes = votos_blancos + votos_nulos + votos_candidaturas,
    turnout = total_votes / censo,
    vote_share_by_party = votes / total_votes
  ) |> 
  ungroup()

# regression model with interaction between turnout and party
model_all <- lm(vote_share_by_party ~ turnout * party_recoded, data = election_data_tidy)

summary(model_all)

# tidy(model_all)


# Let's try to visualise the model in a meaningful way

# Attempt 1

ggplot(election_data_tidy, aes(x = turnout, y = vote_share_by_party, colour = party_recoded)) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Party-Specific Trends: Vote Share vs Turnout",
    x = "Turnout",
    y = "Vote Share",
    colour = "Party"
  ) +
  theme_minimal()


# Attempt 2

data <- election_data_tidy |> 
  mutate(
    turnout_category = cut(
      turnout,
      breaks = c(0, 0.4, 0.6, 0.8, 1), 
      labels = c("Low", "Medium", "High", "Very High"),
      include.lowest = TRUE
    )
  )

summary_data <- data |> 
  group_by(turnout_category, party_recoded) |> 
  summarise(
    avg_vote_share = mean(vote_share_by_party, na.rm = TRUE),
    .groups = "drop"
  )

# Attempt 3
ggplot(summary_data, aes(x = turnout_category, y = avg_vote_share, fill = party_recoded)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Vote Share by Turnout Level and Party",
    x = "Turnout Category",
    y = "Average Vote Share",
    fill = "Party"
  ) +
  theme_minimal()


# Attempt 4 

  ggplot(election_data_tidy, aes(x = party_recoded, y = vote_share_by_party, fill = as.factor(turnout))) +
    geom_violin(trim = FALSE) + 
    labs(
      title = "Vote Share by Party with Voter Turnout",
      x = "Party",
      y = "Vote Share by Party",
      fill = "Turnout"
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) 



```

# Proposal for creative / free questions

Would this be a good idea? Assume 5% of voters arrive after 6 pm. Ask Mafalda

compare differences between the hours and how that benefit the different partıes

maybe some parties get more votes at the beginning

Which means that the predictions over the day favor them first, example: USA republicans are always stronger than democrats in the early hours of the election day.

Could it be that older people vote earlier and young people vote later be a determinant of how the election goes?

The below code addresses the late arrivals (after 6 pm) with a share of 0.05% of whole voters. We might need it so let's keep it passive for now.

```{r, eval=FALSE}
late_arrival_estimate <- 0.05 
adjusted_participacion_2 <- participacion_2 * (1 + late_arrival_estimate)
```

For meeting the filtering criteria for the surveys dataset.

```{r, eval=FALSE}
filtered_surveys <- surveys_tidy |>
  mutate(field_days = as.numeric(difftime(field_date_to, field_date_from, units = "days")) + 1) |>
  filter(year(date_elec) >= 2018,
         is.na(size) | size >= 750,
         field_days >= 1, 
         exit_poll == FALSE)

```

```{r,eval=FALSE}

# We haven't decided yet if we will categorise turnout like this. 

# to decide what makes a low, mid, and high turnout

election_data_tidy <-
  election_data_tidy |> 
  mutate(turnout_category = case_when(
    participacion_2 < 50 ~ "Low Turnout",
    participacion_2 >= 50 & participacion_2 < 70 ~ "Middle Turnout",
    participacion_2 >= 70 ~ "High Turnout"
  ))
```
