---
title: "GROUP TASK"
author: "Pablo Aísa Serranos, Irene Bosque Gala, Diego Fernández Álvarez, Mafalda González González, Sophie Kersten, Irantzu Lamarca Flores, David Pereiro Pol, Gür Piren "
format:
 revealjs:
    theme: [default, style.scss]
    smaller: true
    scrollable: true
    embed-resources: true
    slide-number: true
    code-fold: true
    code-overflow: wrap
execute: 
  echo: true
output: revealjs::revealjs_presentation
---


## Data cleaning

```{r}
#| echo: false
## Libraries 
library(revealjs)
library(tidyverse)
library(readr)
library(glue)
library(ggplot2)
library(ggrepel)
library(lubridate)
```


```{r}
#| eval: false
#| echo: false
election_data <- read_csv(file = "./data/datos_elecciones_brutos.csv")
cod_mun <- read_csv(file = "./data/cod_mun.csv")
surveys <- read_csv(file = "./data/historical_surveys.csv")
abbrev <- read_csv(file = "./data/siglas.csv")
```

```{r}
#| eval: false
#| code-fold: show

important_parties <- c(
  "PARTIDO SOCIALISTA OBRERO ESPAÑOL",
  "PARTIDO POPULAR",
  "CIUDADANOS",
  "EUZKO ALDERDI JELTZALEA-PARTIDO NACIONALISTA VASCO",
  "BLOQUE NACIONALISTA GALEGO",
  "CONVERGÈNCIA I UNIÓ",
  "UNIDAS PODEMOS - IU",
  "ESQUERRA REPUBLICANA DE CATALUNYA",
  "EH - BILDU",
  "MÁS PAÍS",
  "VOX"
)

election_data_tidy <- election_data |> 
  pivot_longer(cols = -(1:15), names_to = "party", values_to = "votes") |> 
  mutate(
    party_recoded = case_when(
      str_detect(party, "PARTIDO SOCIALISTA OBRERO ESPAÑOL|PARTIT DELS SOCIALISTES DE CATALUNYA|PARTIDO SOCIALISTA DE EUSKADI|PARTIDO DOS SOCIALISTAS DE GALICIA") ~ "PARTIDO SOCIALISTA OBRERO ESPAÑOL",
      str_detect(party, "PARTIDO DE LA CIUDADANIA|PARTIDO DE LA CIUDADANÍA") ~ "CIUDADANOS-PARTIDO DE LA CIUDADANIA",
      str_detect(party, "EH - BILDU|ARALAR|ALTERNATIBA|EUSKO ALKARTASUNA") ~ "EUSKAL HERRIA BILDU",
      str_detect(party, "UNIDAS PODEMOS|EN MAREA|PODEM|EZKER BATUA|IZQUIERDA UNIDA|ESQUERRA UNIDA|ESQUERDA UNIDA") ~ "PODEMOS",
      str_detect(party, "CONVERGÈNCIA I UNIÓ|CONVERGENCIA I UNIO|DEMOCRÀCIA I LLIBERTAT|CONVERGÈNCIA i UNIÓ ") ~ "CONVERGENCIA I UNIO",
      str_detect(party, "BLOQUE NACIONALISTA GALEGO|CANDIDATURA GALEGA") ~ "BLOQUE NACIONALISTA GALEGO",
      str_detect(party, "PARTIDO POPULAR") ~ "PARTIDO POPULAR",
      str_detect(party, "MÁS PAÍS") ~ "MÁS PAÍS",
      str_detect(party, "ESQUERRA REPUBLICANA DE CATALUNYA|ESQUERRA REPUBLICANA/CATALUNYA") ~ "ESQUERRA REPUBLICANA DE CATALUNYA",
      party %in% important_parties ~ party,
      TRUE ~ "OTHER"
    ),
    date = glue("{anno}-{mes}-01") |> as_date()
  ) |> 
  unite("cod_mun", codigo_ccaa, codigo_provincia, codigo_municipio, sep = "-", remove = FALSE) |> 
  left_join(
    abbrev |> 
      distinct(denominacion, .keep_all = TRUE) |> 
      mutate(siglas = case_when(
        siglas == "C's" ~ "CS",
        siglas == "EH Bildu" ~ "EH BILDU",
        siglas == "M PAÍS" ~ "MP",
        TRUE ~ siglas
      )), 
    by = c("party_recoded" = "denominacion")
  ) |> 
  left_join(cod_mun, by = "cod_mun") |> 
  select(-vuelta, -tipo_eleccion, -codigo_distrito_electoral) |> 
  drop_na(votes) |> 
  mutate(
    siglas = if_else(is.na(siglas),"OTHER", siglas)
  )

surveys_tidy <- surveys |> 
  pivot_longer(cols = -(1:10), names_to = "party", values_to = "estimation")  |>
  filter(
    year(date_elec) >= 2008,
    exit_poll == FALSE,
    size >= 750,
    field_date_to - field_date_from >= 1
  ) |>
  select(-type_survey) |>
  drop_na(size) 
```

```{r}
#| eval: false
#| echo: false

save(election_data_tidy, file = "data/elections_tidy.rda")
save(surveys_tidy, file = "data/surveys_tidy.rda")

```

```{r}
#| echo: false

load("data/elections_tidy.rda")
load("data/surveys_tidy.rda")
```

--- 

## Question 1
> Which party was the winner in the municipalities with more than 100,000 habitants (census) in each of the elections?

```{r}
# Filter

large_municipalities <- election_data_tidy |>
  filter(censo > 100000)

# Winning parties
winners <- large_municipalities |> 
  group_by(date, municipio) |> 
  slice_max(votes, n = 1, with_ties = FALSE) |> 
  select(date, party_recoded, municipio, censo)

# Number of municipalities won
winners_by_party <- winners |> 
  group_by(date, party_recoded)|> 
  summarize(num_municipalities = n(), .groups = "drop") |> 
  arrange(date, desc(num_municipalities))


```

```{r, fig.height = 8, fig.width = 10}

library(showtext)
font_add_google("Roboto Condensed", "Roboto")
# Modificar el dataframe para que "date" sea un factor con solo las fechas presentes en los datos
winners$date <- factor(winners$date, levels = unique(winners$date))

# Crear el heatmap con los cambios
ggplot(winners, aes(x = date, y = municipio, fill = party_recoded)) +
  geom_tile(color = "white") +
  scale_fill_manual(
    values = c(
      c(
      "PARTIDO POPULAR" = "#1db4e8",
      "PARTIDO SOCIALISTA OBRERO ESPAÑOL" = "#c30505",
      "OTHER" = "gray60",
      "PODEMOS" = "#a444b4",
      "VOX" = "#83b431",
      "EUZKO ALDERDI JELTZALEA-PARTIDO NACIONALISTA VASCO" = "darkgreen",
      "CONVERGENCIA I UNIO" = "#1b348a")
      ),
    labels = c(
      "PARTIDO POPULAR" = "PP",
      "PARTIDO SOCIALISTA OBRERO ESPAÑOL" = "PSOE",
      "OTHER" = "OTHER",
      "PODEMOS" = "PODEMOS",
      "VOX" = "VOX",
      "EUZKO ALDERDI JELTZALEA-PARTIDO NACIONALISTA VASCO" = "PNV",
      "CONVERGENCIA I UNIO" = "CiU"
    )
  ) +
  labs(
    title = "Winning party in municipalities with more than 100,000 habitants",
    x = "Date of election",
    y = "Municipality",
    fill = "Parties"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5, family = "Roboto", margin = margin(b = 20)),
    axis.text.x = element_text(size = 11, family = "Roboto", color = "black"),
    axis.text.y = element_text(size = 11, family = "Roboto", color = "black"),
    legend.title = element_text(size = 11, family = "Roboto", face = "bold"),
    legend.text = element_text(size = 10, family = "Roboto"),
    legend.box.background = element_rect(color = "black", size = 0.5),
    plot.margin = margin(15, 15, 15, 15),
    legend.key.size = unit(1, "lines"),
  ) 


```

--- 

```{r, fig.height = 6, fig.width = 10}

general_election_winners <- data.frame(
  date = as.Date(c("2008-03-01", "2011-11-01", "2015-12-01", "2016-06-01", "2019-04-01", "2019-11-01")),
  party_recoded = c(
    "PARTIDO SOCIALISTA OBRERO ESPAÑOL",  # 2008
    "PARTIDO POPULAR",                    # 2011
    "PARTIDO POPULAR",                    # 2015
    "PARTIDO POPULAR",                    # 2016
    "PARTIDO SOCIALISTA OBRERO ESPAÑOL",  #Primera de 2019
    "PARTIDO SOCIALISTA OBRERO ESPAÑOL")  #Segunda de 2019
)

# Crear la columna end_date (inicio de la siguiente legislatura)
general_election_winners$end_date <- as.Date(c(
  "2011-11-01", "2015-12-01", "2016-06-01", "2019-04-01", "2019-11-01", "2019-11-01"
))

# Gráfico con fondo de color para los partidos ganadores en las elecciones generales
ggplot(winners_by_party, aes(x = date, y = num_municipalities, color = party_recoded)) +
  geom_rect(data = general_election_winners,
            aes(xmin = date, xmax = end_date, ymin = -Inf, ymax = Inf, fill = party_recoded),
            alpha = 0.1, inherit.aes = FALSE) +
  geom_line(size = 1) +
  geom_point(size = 3) +  
  scale_color_manual(
    values = c(
      "PARTIDO POPULAR" = "#1db4e8",
      "PARTIDO SOCIALISTA OBRERO ESPAÑOL" = "#c30505",
      "OTHER" = "gray60",
      "PODEMOS" = "#a444b4",
      "VOX" = "#83b431",
      "EUZKO ALDERDI JELTZALEA-PARTIDO NACIONALISTA VASCO" = "darkgreen",
      "CONVERGENCIA I UNIO" = "#1b348a"
    ),
    labels = c(
      "PARTIDO POPULAR" = "PP",
      "PARTIDO SOCIALISTA OBRERO ESPAÑOL" = "PSOE",
      "OTHER" = "Others",
      "PODEMOS" = "Podemos",
      "VOX" = "Vox",
      "EUZKO ALDERDI JELTZALEA-PARTIDO NACIONALISTA VASCO" = "PNV",
      "CONVERGENCIA I UNIO" = "CiU"
    ))  +
  scale_fill_manual(
    values = c(
      "PARTIDO POPULAR" = "#1db4e8",
      "PARTIDO SOCIALISTA OBRERO ESPAÑOL" = "#c30505"
    ),
    labels = c(
      "PARTIDO POPULAR" = "PP",
      "PARTIDO SOCIALISTA OBRERO ESPAÑOL" = "PSOE")
  ) +
  geom_vline(data = general_election_winners, aes(xintercept = as.numeric(date)),
             color = "gray50", linetype = "dashed", size = 0.4) +
  labs(
    title = "Evolution of winning party in municipalities with more than 100,000 habitants",
    x = "Date of Election",
    y = "Number of Municipalities",
    color = "Winner in each municipality",
    fill = "General Election Winner"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold", family = "Roboto", margin = margin(b = 20)),
    axis.text.x = element_text(size = 11, family = "Roboto", color = "black"),
    axis.text.y = element_text(size = 11, family = "Roboto", color = "black"),
    legend.title = element_text(size = 11, family = "Roboto", face = "bold"),
    legend.text = element_text(size = 10, family = "Roboto"),
    legend.box.background = element_rect(color = "black", size = 0.5),
    plot.margin = margin(15, 15, 15, 15),
    legend.key.size = unit(1, "lines"),
  ) 





```

---

## Question 2

> Which party was the second when the first was the PSOE? And when the first was the PP?

```{r}
ranked_votes <- election_data_tidy |> 
  group_by(date, municipio) |> 
  arrange(desc(votes)) |> 
  mutate(rank = row_number()) |> 
  ungroup()

# PSOE is first
second_psoe <- ranked_votes |> 
  group_by(date, municipio) |> 
  filter(rank == 1 & party_recoded == "PARTIDO SOCIALISTA OBRERO ESPAÑOL") |> 
  left_join(
    ranked_votes |> 
      filter(rank == 2) |> 
      select(date, municipio, second = party_recoded, votes = votes),
    by = c("date", "municipio")
  ) |> 
  ungroup()

second_psoe_sum <- second_psoe |> 
  group_by(date, second) |> 
  summarize(
    num_municipalities = n(),
    .groups = "drop"
  ) |> 
  arrange(date, desc(num_municipalities))

# PP is first
second_pp <- ranked_votes |> 
  group_by(date, municipio) |> 
  filter(rank == 1 & party_recoded == "PARTIDO POPULAR") |> 
  left_join(
    ranked_votes |> 
      filter(rank == 2) |> 
      select(date, municipio, second = party_recoded, votes = votes),
    by = c("date", "municipio")
  ) |> 
  ungroup()

second_pp_sum <- second_pp |> 
  group_by(date, second) |> 
  summarize(
    num_municipalities = n(),
    .groups = "drop"
  ) |> 
  arrange(date, desc(num_municipalities))

```

```{r, fig.height = 6, fig.width = 10}
second_combined <- bind_rows(
  second_pp_sum %>% mutate(first = "PP"),
  second_psoe_sum %>% mutate(first = "PSOE")
)

second_combined$date <- factor(second_combined$date, levels = unique(second_combined$date))

# Gráfico de barras apiladas con facets
ggplot(second_combined, aes(x = date, y = num_municipalities, fill = second)) +
  geom_bar(stat = "identity", position = "fill", color = "black") +
  scale_fill_manual(
    values = c(
      "PARTIDO POPULAR" = "#1db4e8",
      "PARTIDO SOCIALISTA OBRERO ESPAÑOL" = "#c30505",
      "OTHER" = "gray60",
      "PODEMOS" = "#a444b4",
      "EUZKO ALDERDI JELTZALEA-PARTIDO NACIONALISTA VASCO" = "darkgreen",
      "BLOQUE NACIONALISTA GALEGO" = "lightblue",
      "CIUDADANOS-PARTIDO DE LA CIUDADANIA" = "orange",
      "VOX" = "#83b431",
      "CONVERGENCIA I UNIO" = "#1b348a",
      "ESQUERRA REPUBLICANA DE CATALUNYA"= "yellow"
    ),
    labels = c(
      "PARTIDO POPULAR" = "PP",
      "PARTIDO SOCIALISTA OBRERO ESPAÑOL" = "PSOE",
      "OTHER" = "Others",
      "PODEMOS" = "Podemos",
      "EUZKO ALDERDI JELTZALEA-PARTIDO NACIONALISTA VASCO" = "PNV",
      "BLOQUE NACIONALISTA GALEGO" = "BNG",
      "CIUDADANOS-PARTIDO DE LA CIUDADANIA" = "C's",
      "CONVERGENCIA I UNIO" = "CiU",
      "ESQUERRA REPUBLICANA DE CATALUNYA"= "ERC"
    )
  ) +
  labs(
    title = "Second places when PSOE or PP were first",
    x = "Election Date",
    y = "Number of Municipalities",
    fill = "Second Party"
  ) +
  facet_wrap(~ first, scales = "free_y", labeller = labeller(first = c(PP = "PP First", PSOE = "PSOE First"))) +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 14, face = "bold", family = "Roboto"),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5, family = "Roboto", margin = margin(b = 20)),
    axis.text.x = element_text(size = 11, family = "Roboto", color = "black", angle = 20),
    axis.text.y = element_text(size = 11, family = "Roboto", color = "black"),
    legend.title = element_text(size = 11, family = "Roboto", face = "bold"),
    legend.text = element_text(size = 10, family = "Roboto"),
    legend.box.background = element_rect(color = "black", size = 0.5),
    plot.margin = margin(15, 15, 15, 15),
    legend.key.size = unit(1, "lines")
  ) 

```

---

```{r, fig.height = 7, fig.width = 12}
library(ggalluvial)

# Recodificar los nombres de los partidos
second_combined <- second_combined %>%
  mutate(
    first = recode(first,
      "PARTIDO POPULAR" = "PP",
      "PARTIDO SOCIALISTA OBRERO ESPAÑOL" = "PSOE",
      "OTHER" = "Others",
      "PODEMOS" = "Podemos",
      "EUZKO ALDERDI JELTZALEA-PARTIDO NACIONALISTA VASCO" = "PNV",
      "BLOQUE NACIONALISTA GALEGO" = "BNG",
      "CIUDADANOS-PARTIDO DE LA CIUDADANIA" = "C's",
      "CONVERGENCIA I UNIO" = "CiU",
      "ESQUERRA REPUBLICANA DE CATALUNYA" = "ERC"
    ),
    second = recode(second,
      "PARTIDO POPULAR" = "PP",
      "PARTIDO SOCIALISTA OBRERO ESPAÑOL" = "PSOE",
      "OTHER" = "Others",
      "PODEMOS" = "Podemos",
      "EUZKO ALDERDI JELTZALEA-PARTIDO NACIONALISTA VASCO" = "PNV",
      "BLOQUE NACIONALISTA GALEGO" = "BNG",
      "CIUDADANOS-PARTIDO DE LA CIUDADANIA" = "C's",
      "CONVERGENCIA I UNIO" = "CiU",
      "ESQUERRA REPUBLICANA DE CATALUNYA" = "ERC"
    )) |> 
  filter(!(second == "PNV" | second == "BNG" | second == "ERC"))
  

# Gráfico Alluvial actualizado
ggplot(second_combined, aes(
  axis1 = first, axis2 = second,
  y = num_municipalities, fill = second
)) +
  geom_alluvium(aes(fill = second), width = 1/6) +
  geom_stratum(aes(fill = after_stat(stratum)), width = 1/6, color = "black") +
 geom_text(stat = "stratum", aes(label = after_stat(stratum)), size = 3.5, color = "black", fontface = "bold") +
  scale_fill_manual(
    values = c(
      "PP" = "#1db4e8",
      "PSOE" = "#c30505",
      "Others" = "gray60",
      "Podemos" = "#a444b4",
      "PNV" = "darkgreen",
      "BNG" = "lightblue",
      "C's" = "orange",
      "VOX" = "#83b431",
      "CiU" = "#1b348a",
      "ERC" = "yellow"
    )
  ) +
  labs(
    title = "Flow of municipalities won: First to second party",
    x = "First to Second Party",
    y = "Number of Municipalities",
    fill = "Second Party"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 17, face = "bold", hjust = 0.5, family = "Roboto", margin = margin(b = 20)),
    axis.text.x = element_text(size = 12, family = "Roboto", color = "black"),
    axis.text.y = element_text(size = 12, family = "Roboto", color = "black"),
    axis.title.x = element_text(size = 13),
    axis.title.y = element_text(size = 13),
    legend.title = element_text(size = 13, family = "Roboto", face = "bold"),
    legend.text = element_text(size = 11, family = "Roboto"),
    legend.box.background = element_rect(color = "black", size = 0.5),
    plot.margin = margin(15, 15, 15, 15),
    legend.key.size = unit(1, "lines")
  )
```

---

## Question 3

> Who benefits from low turnout?

```{r}

election_data_tidy <- election_data_tidy |> 
  group_by(cod_mun, date, party_recoded) |> 
  mutate(
    total_votes = votos_blancos + votos_nulos + votos_candidaturas,
    turnout = total_votes / censo,
    votes_recoded = sum(votes, na.rm = TRUE),
    vote_share_by_party = votes_recoded / total_votes
  ) |> 
  ungroup()

# Let's try to visualise the model in a meaningful way


ggplot(election_data_tidy, aes(x = turnout, y = vote_share_by_party, colour = siglas)) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Party-Specific Trends: Vote Share vs Turnout",
    x = "Turnout",
    y = "Vote Share",
    colour = "Party"
  ) +
  theme_minimal() +
  scale_colour_manual(values = c(
      "PP" = "#1db4e8",
      "PSOE" = "#c30505",
      "OTHER" = "gray60",
      "PODEMOS" = "#a444b4",
      "VOX" = "#83b431",
      "ERC" = "#ffbf41",
      "CIU" = "#1b348a",
      "MP" = "#004938",
      "CS" = "#eb6109",
      "EAJ-PNV" = "darkgreen",
      "BNG" = "lightblue",
      "EH BILDU" = "#03cfb4"
    )) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 17, face = "bold", hjust = 0.5, family = "Roboto", margin = margin(b = 20)),
    axis.text.x = element_text(size = 12, family = "Roboto", color = "black"),
    axis.text.y = element_text(size = 12, family = "Roboto", color = "black"),
    axis.title.x = element_text(size = 13),
    axis.title.y = element_text(size = 13),
    legend.title = element_text(size = 13, family = "Roboto", face = "bold"),
    legend.text = element_text(size = 11, family = "Roboto"),
    legend.box.background = element_rect(color = "black", size = 0.5),
    plot.margin = margin(15, 15, 15, 15),
    legend.key.size = unit(1, "lines")
  )

```

---

## Question 4

> How to analyze the relationship between census and vote? 

[@gómezvalenzuela2023]

- Rural < 10000 recorded citizens in the census
- Urban > 10000 recorded citizens in the census


```{r}

ggplot(election_data_tidy, aes(x = censo, y = vote_share_by_party, colour = siglas)) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Party-Specific Trends: Vote Share vs Census",
    x = "Census",
    y = "Vote Share",
    colour = "Party"
  ) +
  theme_minimal() +
  scale_colour_manual(values = c(
      "PP" = "#1db4e8",
      "PSOE" = "#c30505",
      "OTHER" = "gray60",
      "PODEMOS" = "#a444b4",
      "VOX" = "#83b431",
      "ERC" = "#ffbf41",
      "CIU" = "#1b348a",
      "MP" = "#004938",
      "CS" = "#eb6109",
      "EAJ-PNV" = "darkgreen",
      "BNG" = "lightblue",
      "EH BILDU" = "#03cfb4"
    )) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 17, face = "bold", hjust = 0.5, family = "Roboto", margin = margin(b = 20)),
    axis.text.x = element_text(size = 12, family = "Roboto", color = "black"),
    axis.text.y = element_text(size = 12, family = "Roboto", color = "black"),
    axis.title.x = element_text(size = 13),
    axis.title.y = element_text(size = 13),
    legend.title = element_text(size = 13, family = "Roboto", face = "bold"),
    legend.text = element_text(size = 11, family = "Roboto"),
    legend.box.background = element_rect(color = "black", size = 0.5),
    plot.margin = margin(15, 15, 15, 15),
    legend.key.size = unit(1, "lines")
  )
```

---

> Is it true that certain parties win in rural areas?

```{r}
# For the second part - rural vs urban

election_data_tidy <- election_data_tidy |> 
  mutate(
    area_type = ifelse(censo < 10000, "Rural", "Urban")
  )
```


```{r}
#| echo: false
#| eval: false

# Statistical information
model_area_type <- lm(vote_share_by_party ~ area_type * party_recoded, data = election_data_tidy)
summary(model_area_type)
```


```{r}
rural_municipalities <- election_data_tidy |>
  filter(area_type == "Rural")

# Winning parties
winners_rural <- rural_municipalities |> 
  group_by(date, municipio) |> 
  slice_max(votes_recoded, n = 1, with_ties = FALSE) |> 
  select(date, party_recoded, municipio, censo)

# Number of municipalities won
winners_by_party_rural <- winners |> 
  group_by(date, party_recoded)|> 
  summarize(num_municipalities = n(), .groups = "drop") |> 
  arrange(date, desc(num_municipalities))|> 
  ungroup() 

urban_municipalities <- election_data_tidy |>
  filter(area_type == "Urban")

# Winning parties
winners_urban <- urban_municipalities |> 
  group_by(date, municipio) |> 
  slice_max(votes_recoded, n = 1, with_ties = FALSE) |> 
  select(date, party_recoded, municipio, censo)

# Number of municipalities won
winners_by_party_urban <- winners_urban |> 
  group_by(date, party_recoded)|> 
  summarize(num_municipalities = n(), .groups = "drop") |> 
  arrange(date, desc(num_municipalities))|> 
  ungroup() 

type_combined <- rbind(
  winners_by_party_rural %>% mutate(type = "Rural"),
  winners_by_party_urban %>% mutate(type = "Urban")
)

type_combined$date <- factor(type_combined$date, levels = unique(type_combined$date))

ggplot(type_combined, aes(x = date, y = num_municipalities, fill = party_recoded)) +
  geom_bar(stat = "identity", position = "fill", color = "black") +
  scale_fill_manual(
    values = c(
      "PARTIDO POPULAR" = "#1db4e8",
      "PARTIDO SOCIALISTA OBRERO ESPAÑOL" = "#c30505",
      "OTHER" = "gray60",
      "PODEMOS" = "#a444b4",
      "EUZKO ALDERDI JELTZALEA-PARTIDO NACIONALISTA VASCO" = "darkgreen",
      "BLOQUE NACIONALISTA GALEGO" = "lightblue",
      "CIUDADANOS-PARTIDO DE LA CIUDADANIA" = "orange",
      "VOX" = "#83b431",
      "CONVERGENCIA I UNIO" = "#1b348a",
      "ESQUERRA REPUBLICANA DE CATALUNYA"= "yellow"
    ),
    labels = c(
      "PARTIDO POPULAR" = "PP",
      "PARTIDO SOCIALISTA OBRERO ESPAÑOL" = "PSOE",
      "OTHER" = "Others",
      "PODEMOS" = "Podemos",
      "EUZKO ALDERDI JELTZALEA-PARTIDO NACIONALISTA VASCO" = "PNV",
      "BLOQUE NACIONALISTA GALEGO" = "BNG",
      "CIUDADANOS-PARTIDO DE LA CIUDADANIA" = "C's",
      "CONVERGENCIA I UNIO" = "CiU",
      "ESQUERRA REPUBLICANA DE CATALUNYA"= "ERC"
    )
  ) +
  labs(
    title = "Wins depending on the type of the area",
    x = "Election Date",
    y = "Number of Municipalities",
    fill = "Parties"
  ) +
  facet_wrap(~ type, scales = "free_y") +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 14, face = "bold", family = "Roboto"),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5, family = "Roboto", margin = margin(b = 20)),
    axis.text.x = element_text(size = 11, family = "Roboto", color = "black", angle = 20),
    axis.text.y = element_text(size = 11, family = "Roboto", color = "black"),
    legend.title = element_text(size = 11, family = "Roboto", face = "bold"),
    legend.text = element_text(size = 10, family = "Roboto"),
    legend.box.background = element_rect(color = "black", size = 0.5),
    plot.margin = margin(15, 15, 15, 15),
    legend.key.size = unit(1, "lines")
  ) 

```

---

## Question 5

> How to calibrate the error of the polls (remember that the polls are voting intentions at national level)?

```{r}

elections_aggregated2 <- election_data_tidy |> 
  group_by(date, cod_mun) |> 
  distinct(total_votes, .keep_all = TRUE) |> 
  summarize(
    participation = sum(total_votes),
    .groups = "drop"
  ) |> 
  group_by(date) |> 
  summarise(participation_total = sum(participation),
    .groups = "drop"
  )

elections_aggregated1 <- election_data_tidy |> 
  group_by(date, party_recoded, siglas) |> 
  summarize(
    total_votes_all = sum(votes, na.rm = TRUE),
    .groups = "drop"
  ) 

elections_with_shares <- elections_aggregated1 |> 
  left_join(elections_aggregated2, by = "date") |> 
  mutate(vote_share = (total_votes_all / participation_total) * 100)

```

```{r}

poll_calibration <- surveys_tidy |> 
  mutate(year_month_elec = floor_date(date_elec, "month")) |> # I will extract year and month (lubridate package)
  left_join(
    elections_with_shares |> 
      mutate(year_month = floor_date(date, "month")),          # I will extract year and month
    by = c("year_month_elec" = "year_month", "party" = "siglas")
  )

poll_calibration <- poll_calibration |> 
  mutate(error = estimation - vote_share)

poll_calibration |> select(error)

```

---

## Question 6

> Which polling houses got it right the most and which ones deviated the most from the results?

```{r}

# Error analysis: summary of errors by pollster or any other factor

error_analysis <- poll_calibration |> 
  group_by(pollster) |> 
  summarize(
    mean_error = mean(abs(error), na.rm = TRUE),
    sd_error = sd(error, na.rm = TRUE)
  )

```

```{r}

poll_calibration <- poll_calibration |> 
  mutate(abs_error = abs(error))  

pollster_accuracy <- poll_calibration |> 
  group_by(pollster) |> 
  summarize(
    mean_abs_error = mean(abs_error, na.rm = TRUE),# mean abs error column
    sd_abs_error = sd(abs_error, na.rm = TRUE)
  ) |> 
  arrange(mean_abs_error)

# Bar plot to represent the MAE for the hightst 10 pollsters
top_pollsters <- pollster_accuracy %>%
  slice_max(mean_abs_error, n = 10)  

ggplot(top_pollsters, aes(x = reorder(pollster, mean_abs_error), y = mean_abs_error)) +
  geom_bar(stat = "identity", fill = "purple", alpha = 0.7) +
  labs(
    title = "Top 10 Pollster with Highest Mean Absolute Error",
    x = "Pollster",
    y = "Mean Absolute Error"
  ) +
  coord_flip() +
  theme_minimal()

```

---

```{r}
# Bar plot to represent the MAE for the lowest 10 pollsters
low_pollsters <- pollster_accuracy %>%
  slice_min(mean_abs_error, n = 10)  

ggplot(low_pollsters, aes(x = reorder(pollster, mean_abs_error), y = mean_abs_error)) +
  geom_bar(stat = "identity", fill = "purple", alpha = 0.7) +
  labs(
    title = "Top 10 Pollster with Lowest Mean Absolute Error",
    x = "Pollster",
    y = "Mean Absolute Error"
  ) +
  coord_flip() +
  theme_minimal()
```

---

## Creative 1

> How does polling error vary by party?

```{r}

polling_error_by_party <- poll_calibration |> 
  group_by(party) |> 
  summarize(
    mean_abs_error = mean(abs_error, na.rm = TRUE),
    sd_abs_error = sd(abs_error, na.rm = TRUE)
  ) |> 
  arrange(mean_abs_error) |> 
  head(10)

polling_error_by_party 

```
---

## Creative 2

---

## Creative 3

---

## Creative 4








